---
title: "BioPRIO"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# BioPRIO -- Biological Invasive species Prioritization

A Shiny tool for assessing the invasion risk of **terrestrial invertebrates** in Norway. Adapted from the [FinnPRIO](https://doi.org/10.1007/s10530-016-1126-3) plant-pest framework, using generic "risk assessment area" terminology.

## Assessment Framework

Monte Carlo simulation (default 50,000 iterations) with PERT distributions across four modules:

| Module | What it scores |
|--------|----------------|
| **ENT** (Entry) | Probability of entering via pathways |
| **EST** (Establishment) | Likelihood of establishing and spreading |
| **IMP** (Impact) | Economic and environmental/social impact |
| **MAN** (Management) | Preventability and controllability |

`RISK = IMPACT × INVASION`, where `INVASION = ENTRY × ESTABLISHMENT` and `MANAGEABILITY = min(PREVENTABILITY, CONTROLLABILITY)`.

## Quick Start

Requires **R ≥ 4.0** (packages auto-install on first run via `global.R`).

```{r}
shiny::runApp()        # or: source("START_APP.R")
```

On startup, select a SQLite `.db` file containing species and assessment data.

## AI-Assisted Justifications (optional)

Python scripts (≥ 3.9) generate scientific justifications and values for assessment answers. API keys (OpenAI, Tavily) are read from external files, never the repo.

```{bash}
# 1. Fetch literature (also: scripts/get litterature/get_species_literature.R)
python python/support/get_additional_literature.py

# 2. Generate justifications (web + local PDFs)
python python/gpt_researcher_scripts/populate_bioprio_justifications_hybrid.py

# 3. Determine min/likely/max values
python python/gpt_researcher_scripts/populate_bioprio_values.py
```

## Database Population

```{r}
# Invertebrates (GBIF taxonomy + distribution)
source("scripts/populate database/populate_ant_species.R")
```

Original FinnPRIO/EPPO scripts live in `scripts/populate database scripts/`.

## Project Layout

| Path | Contents |
|------|----------|
| `R/` | Core modules (`simulations.R` holds the scoring logic) |
| `ui.R`, `server.R`, `global.R` | Shiny app |
| `python/` | AI justification pipelines (GPT Researcher, SQuAI) |
| `scripts/` | Literature fetching, database population, migration |
| `www/` | Web assets and user guide |

## Terminology (BioPRIO vs FinnPRIO)

| FinnPRIO | BioPRIO |
|----------|---------|
| pest | species |
| PRA area | risk assessment area |
| host plant | suitable host, prey, or habitat |

## References

- Heikkilä et al. (2016) *A novel prioritizing method for invasive alien plant pests.* Biological Invasions 18:1827--1842. [DOI: 10.1007/s10530-016-1126-3](https://doi.org/10.1007/s10530-016-1126-3)
- [CBD Pathways of Introduction](https://www.cbd.int/doc/decisions/cop-12/cop-12-dec-17-en.pdf)

## License

Internal use -- Norwegian Scientific Committee for Food and Environment (VKM).
